---
title: "Practical Homework-2"
author: "Sravya Murala"
date: "2026-04-25"
output: html_document
---

## Introduction

This homework is worked at predicting whether a person has ever been diagnosed with cancer using their demographics and lifestyle behaviors. The data is from the 2022 National Health Interview Survey (NHIS), accessed through IPUMS Health Surveys. I tune three SVM kernels the linear, radial, and polynomial using 5-fold cross-validation, then compare them on a held-out test set. The goal is to see whether non-linear kernels capture more signal than a simple linear one, and to identify which variables matter most for cancer prediction.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(e1071)
library(caret)
library(dplyr)
library(pROC)
library(ggplot2)
library(gridExtra)
library(magrittr)
library(kernlab)
# Parallel processing to use multiple CPU cores
library(doParallel)
cl <- makePSOCKcluster(parallel::detectCores() - 1)  # use all cores except 1
registerDoParallel(cl)
```

#### Loading and Data Inspection

```{r}
# Loading the Data
nhis_df <- read.csv("/Users/sravyaaas/Documents/SML-2/practical homework-2/nhis_2022.csv")

# Checking dimensions and structure
dim(nhis_df)
glimpse(nhis_df)
```

```{r}
table(nhis_df$CANCEREV, useNA = "ifany")
```

```{r}
# selecting Variables, confirming variables planned to use as predictors
my_vars <- c("AGE", "SEX", "POVERTY", "BMICALC",
             "MOD10DMIN", "HRSLEEP", "ALCDAYSYR", "FRUTNO")

all(my_vars %in% names(nhis_df))

summary(nhis_df[, my_vars])
```

#### Data Cleaning

The NHIS uses sentinel codes 996, 997, 998, 999, etc. for missing,refused, or "Not in Universe" responses. These must be converted to NA before modeling, otherwise the SVM will treat "999" as a real value.

```{r}
# Filtering adults
nhis_clean <- subset(nhis_df, is.element(CANCEREV, c(1, 2)) & AGE >= 18 & AGE <= 85)
dim(nhis_clean)
```

```{r}
# Recoding target
nhis_clean$CANCEREV <- factor(ifelse(nhis_clean$CANCEREV == 2, "Yes", "No"),
                              levels = c("No", "Yes"))
table(nhis_clean$CANCEREV)
prop.table(table(nhis_clean$CANCEREV))
```

```{r}
# Recoding and dropping missing values
# Keeping only columns needed for modeling
cols <- c("CANCEREV", "AGE", "SEX", "POVERTY", "BMICALC",
               "MOD10DMIN", "HRSLEEP", "ALCDAYSYR", "FRUTNO")
nhis_clean <- nhis_clean[, cols]
```

```{r}
# Converting to a factor as SVM treats numeric 1/2 as ordered which is wrong
nhis_clean$SEX[is.element(nhis_clean$SEX, c(7, 8, 9))] <- NA
nhis_clean$SEX <- factor(ifelse(nhis_clean$SEX == 1, "Male", "Female"),
                         levels = c("Male", "Female"))
```

```{r}
# 98 = Undefinable, 99 = Unknown -> NA
nhis_clean$POVERTY[is.element(nhis_clean$POVERTY, c(98, 99))] <- NA
nhis_clean$BMICALC[nhis_clean$BMICALC == 0 | nhis_clean$BMICALC >= 99] <- NA
nhis_clean$MOD10DMIN[nhis_clean$MOD10DMIN >= 996] <- NA

nhis_clean$HRSLEEP[nhis_clean$HRSLEEP == 25] <- 0.5
nhis_clean$HRSLEEP[is.element(nhis_clean$HRSLEEP, c(0, 97, 98, 99))] <- NA

nhis_clean$ALCDAYSYR[nhis_clean$ALCDAYSYR >= 995] <- NA

nhis_clean$FRUTNO[nhis_clean$FRUTNO >= 995] <- NA
```


```{r}
# Dropping rows with NAs in any predictor
nhis_clean <- na.omit(nhis_clean)

# Verify the cleaned data
dim(nhis_clean)
summary(nhis_clean)
```

```{r}
# checking for class balance 
table(nhis_clean$CANCEREV)
round(prop.table(table(nhis_clean$CANCEREV)) * 100, 2)
```

###  Exploratory Data Analysis

Before modeling, Its important that we look at how the target relates to a few key predictors. This helps confirm that the cleaned data behaves the way we would expect and gives a sense of which variables might carry signal.

```{r eda-plots, fig.width=12, fig.height=4}
# Age by cancer status 
p1 <- ggplot(nhis_clean, aes(x = CANCEREV, y = AGE, fill = CANCEREV)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("No" = "#1f77b4", "Yes" = "#d62728")) +
  labs(title = "Age by Cancer Status", x = "Ever Had Cancer", y = "Age (years)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

# BMI by cancer status
p2 <- ggplot(nhis_clean, aes(x = CANCEREV, y = BMICALC, fill = CANCEREV)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("No" = "#1f77b4", "Yes" = "#d62728")) +
  labs(title = "BMI by Cancer Status", x = "Ever Had Cancer", y = "BMI") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

# Cancer rate by sex
sex_summary <- aggregate(rep(1, nrow(nhis_clean)),
                         by = list(SEX = nhis_clean$SEX,
                                   CANCEREV = nhis_clean$CANCEREV),
                         FUN = sum)
names(sex_summary)[3] <- "n"
sex_summary$prop <- ave(sex_summary$n, sex_summary$SEX, FUN = function(x) x / sum(x))

p3 <- ggplot(sex_summary, aes(x = SEX, y = prop, fill = CANCEREV)) +
  geom_col(position = "dodge", alpha = 0.85) +
  scale_fill_manual(values = c("No" = "#1f77b4", "Yes" = "#d62728")) +
  labs(title = "Cancer Rate by Sex", x = "Sex", y = "Proportion", fill = "Cancer") +
  theme_minimal(base_size = 12)

gridExtra::grid.arrange(p1, p2, p3, nrow = 1)

```

#### Data preparation

```{r}
# Set seed for reproducibility 
set.seed(2025)

# Stratified subsample
sample_strat <- caret::createDataPartition(nhis_clean$CANCEREV, 
                                         p = 6000 / nrow(nhis_clean), 
                                         list = FALSE)
nhis_sample <- nhis_clean[sample_strat, ]

# verifying sample size 
dim(nhis_sample)
round(prop.table(table(nhis_sample$CANCEREV)) * 100, 2)
```

```{r}
# Splitting train and test
train_lists <- caret::createDataPartition(nhis_sample$CANCEREV, 
                                        p = 0.8, 
                                        list = FALSE)
train_df <- nhis_sample[train_lists, ]
test_df  <- nhis_sample[-train_lists, ]

# Checking sizes and class balance 
cat("Train set:\n")
dim(train_df)
table(train_df$CANCEREV)

cat("\nTest set:\n")
dim(test_df)
table(test_df$CANCEREV)
```

As we know SVMs are distance-based, so features on different scales would be unfair the AGE 0-85 would dominate HRSLEEP 1-24. So we scale to mean=0, sd=1. We compute scaling parameters on the TRAINING set only, then apply those same parameters to the test set to avoid any data leakage.

```{r}
# scaling numeric columns
numeric_columns <- c("AGE", "POVERTY", "BMICALC", "MOD10DMIN", 
                  "HRSLEEP", "ALCDAYSYR", "FRUTNO")

# scaling values
preprocessing <- caret::preProcess(train_df[, numeric_columns], 
                             method = c("center", "scale"))

train_df[, numeric_columns] <- predict(preprocessing, train_df[, numeric_columns])
test_df[, numeric_columns]  <- predict(preprocessing, test_df[, numeric_columns])

summary(train_df[, numeric_columns])
```

#### Cross validation

```{r}
# CV control
cv_config <- caret::trainControl(method = "cv",
                                 number = 5, 
                                 classProbs = TRUE, 
                                 summaryFunction = caret::twoClassSummary,
                                 savePredictions = "final",
                                 sampling = NULL) 
```

## Modeling

I am tunining three SVM kernels separately using 5-fold cross-validation, scoring each by ROC-AUC because of the class imbalance of ~13% cancer. Each kernel has its own hyperparameter grid which the linear has just C, radial adds sigma, and polynomial adds degree and scale. Class weights are passed to the SVM to address the imbalance without resampling.

#### Linear SVM Tuning

```{r}
# Linear kernal
grid_linearSVM <- expand.grid(C = 10^seq(-2, 2, length.out = 9))

# Class weights
class_weights <- c(No = 1, Yes = sum(train_df$CANCEREV == "No") / sum(train_df$CANCEREV == "Yes"))

set.seed(2025) 
svm_linear <- caret::train(CANCEREV ~ ., 
                           data          = train_df,
                           method        = "svmLinear",
                           trControl     = cv_config,
                           tuneGrid      = grid_linearSVM,
                           metric        = "ROC",
                           class.weights = class_weights)

cat("Best C for linear kernel:\n")
print(svm_linear$bestTune)

cat("\nFull CV results:\n")
print(svm_linear$results)
```

#### RBF- Radial SVM Tuning

```{r}
# RBF adds sigma to tuning
grid_radial <- expand.grid(C     = 10^seq(-1, 2, length.out = 5),
                           sigma = 10^seq(-3, 0, length.out = 5))

set.seed(2025) 
svm_radial <- caret::train(CANCEREV ~ ., 
                           data          = train_df,
                           method        = "svmRadial",
                           trControl     = cv_config,
                           tuneGrid      = grid_radial,
                           metric        = "ROC",
                           class.weights = class_weights)

cat("Best parameters for radial kernel:\n")
print(svm_radial$bestTune)

cat("\nFull CV results:\n")
print(svm_radial$results)
```

#### Polynomial SVM Tuning

```{r}
# Polynomial kernal
grid_poly <- expand.grid(C      = 10^seq(-1, 1, length.out = 3),
                         degree = c(2, 3),
                         scale  = c(0.01, 0.1, 1))

set.seed(2025) 
svm_poly <- caret::train(CANCEREV ~ ., 
                         data          = train_df,
                         method        = "svmPoly",
                         trControl     = cv_config,
                         tuneGrid      = grid_poly,
                         metric        = "ROC",
                         class.weights = class_weights)

cat("Best parameters for polynomial kernel:\n")
print(svm_poly$bestTune)

cat("\nFull CV results:\n")
print(svm_poly$results)
```

#### Evaluation

```{r}
evaluate_svm <- function(model, test_data, model_name) {
  
  # Getting predicted probabilities 
  predicted_prob <- predict(model, test_data, type = "prob")[, "Yes"]
  
  # ROC and AUC 
  roc_curve <- pROC::roc(response  = test_data$CANCEREV,
                       predictor = predicted_prob,
                       levels    = c("No", "Yes"),
                       quiet     = TRUE)
  auc_value <- as.numeric(pROC::auc(roc_curve))
  
  # optimal probability threshold
  thresholds <- seq(0.05, 0.95, by = 0.01)
  f1_scores <- sapply(thresholds, function(t) {
    pred_class <- factor(ifelse(predicted_prob >= t, "Yes", "No"),
                         levels = c("No", "Yes"))
    confmatrix <- caret::confusionMatrix(pred_class, test_data$CANCEREV,
                                 positive = "Yes", mode = "everything")
    f1 <- confmatrix$byClass["F1"]
    ifelse(is.na(f1), 0, f1)
  })
  best_threshold <- thresholds[which.max(f1_scores)]
  
  # confusion matrix at best threshold
  pred_class <- factor(ifelse(predicted_prob >= best_threshold, "Yes", "No"),
                       levels = c("No", "Yes"))
  confmatrix <- caret::confusionMatrix(pred_class, test_data$CANCEREV,
                               positive = "Yes", mode = "everything")
  
  list(
    name       = model_name,
    threshold  = best_threshold,
    accuracy   = unname(confmatrix$overall["Accuracy"]),
    precision  = unname(confmatrix$byClass["Precision"]),
    recall     = unname(confmatrix$byClass["Recall"]),
    f1         = unname(confmatrix$byClass["F1"]),
    auc        = auc_value,
    confusion  = confmatrix$table,
    roc        = roc_curve,
    predicted_prob  = predicted_prob
  )
}
```

```{r}
# Evaulating three kernals

results_linear <- evaluate_svm(svm_linear, test_df, "Linear")
results_radial <- evaluate_svm(svm_radial, test_df, "Radial (RBF)")
results_poly   <- evaluate_svm(svm_poly,   test_df, "Polynomial")
```

```{r}
# Comparison 
results_table <- data.frame(
  Model     = c(results_linear$name, results_radial$name, results_poly$name),
  Threshold = round(c(results_linear$threshold, results_radial$threshold, results_poly$threshold), 2),
  Accuracy  = round(c(results_linear$accuracy,  results_radial$accuracy,  results_poly$accuracy),  3),
  Precision = round(c(results_linear$precision, results_radial$precision, results_poly$precision), 3),
  Recall    = round(c(results_linear$recall,    results_radial$recall,    results_poly$recall),    3),
  F1        = round(c(results_linear$f1,        results_radial$f1,        results_poly$f1),        3),
  ROC_AUC   = round(c(results_linear$auc,       results_radial$auc,       results_poly$auc),       3)
)

print(results_table)
results_table$Model <- c("Linear SVM", "Radial (RBF) SVM", "Polynomial SVM")
print(results_table)
```

All three kernels perform similarly with AUC around 0.77-0.78, with radial slightly ahead. The recall is around 70% across kernels, which means the model catches about 7 out of 10 cancer cases, but precision is low which is ~27% because of the low base rate of cancer in the data. The confusion matrices below show this in raw counts.

```{r}
# Confusion matrices
cat("\n Linear SVM confusion matrix \n")
print(results_linear$confusion)
cat("\n Radial SVM confusion matrix \n")
print(results_radial$confusion)
cat("\n Polynomial SVM confusion matrix \n")
print(results_poly$confusion)
```

```{r}
# Combining ROC curves 
roc_data <- rbind(
  data.frame(FPR = 1 - results_linear$roc$specificities,
             TPR = results_linear$roc$sensitivities,
             Model = "Linear"),
  data.frame(FPR = 1 - results_radial$roc$specificities,
             TPR = results_radial$roc$sensitivities,
             Model = "Radial (RBF)"),
  data.frame(FPR = 1 - results_poly$roc$specificities,
             TPR = results_poly$roc$sensitivities,
             Model = "Polynomial")
)

# AUC labels 
auc_labels <- paste0(c("Linear", "Radial (RBF)", "Polynomial"),
                     " (AUC = ",
                     round(c(results_linear$auc, results_radial$auc, results_poly$auc), 3),
                     ")")
roc_data$Model <- factor(roc_data$Model,
                         levels = c("Linear", "Radial (RBF)", "Polynomial"),
                         labels = auc_labels)

roc_plot <- ggplot(roc_data, aes(x = FPR, y = TPR, color = Model)) +
  geom_line(size = 1.1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60") +
  labs(title    = "ROC Curves: SVM Kernels for Cancer Prediction",
       subtitle = "Test set performance, n = 1,199",
       x        = "False Positive Rate (1 - Specificity)",
       y        = "True Positive Rate (Sensitivity)",
       color    = NULL) +
  scale_color_manual(values = c("#1f77b4", "#d62728", "#2ca02c")) +
  theme_minimal(base_size = 13) +
  theme(legend.position = c(0.65, 0.2),
        legend.background = element_rect(fill = "white", color = "grey80"))

print(roc_plot)

ggsave("roc_curves.png", roc_plot, width = 8, height = 6, dpi = 300)
```

```{r}
# confusion matrix heatmap
make_cm_plot <- function(cm_table, title) {
  cm_df <- as.data.frame(cm_table)
  ggplot(cm_df, aes(x = Reference, y = Prediction, fill = Freq)) +
    geom_tile(color = "white") +
    geom_text(aes(label = Freq), color = "white", size = 6, fontface = "bold") +
    scale_fill_gradient(low = "#9ecae1", high = "#08306b") +
    labs(title = title, x = "Actual", y = "Predicted") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, face = "bold"))
}

cm_plot_linear <- make_cm_plot(results_linear$confusion, "Linear SVM")
cm_plot_radial <- make_cm_plot(results_radial$confusion, "Radial (RBF) SVM")
cm_plot_poly   <- make_cm_plot(results_poly$confusion,   "Polynomial SVM")

cm_combined <- gridExtra::grid.arrange(cm_plot_linear, cm_plot_radial, cm_plot_poly,
                                       nrow = 1)

ggsave("confusion_matrices.png", cm_combined, width = 12, height = 4, dpi = 300)
```

```{r}
# Permutation importance

permutation_importance <- function(model, data, target_col = "CANCEREV") {
  base_pred <- predict(model, data, type = "prob")[, "Yes"]
  base_auc <- as.numeric(pROC::auc(pROC::roc(data[[target_col]], base_pred,
                                             levels = c("No", "Yes"), quiet = TRUE)))
  
  features <- setdiff(names(data), target_col)
  importance <- numeric(length(features))
  names(importance) <- features
  
  for (f in features) {
    permuted <- data
    permuted[[f]] <- sample(permuted[[f]])  # shuffle this column
    new_pred <- predict(model, permuted, type = "prob")[, "Yes"]
    new_auc <- as.numeric(pROC::auc(pROC::roc(permuted[[target_col]], new_pred,
                                              levels = c("No", "Yes"), quiet = TRUE)))
    importance[f] <- base_auc - new_auc
  }
  importance
}

set.seed(2025)
imp_radial <- permutation_importance(svm_radial, test_df)

imp_df <- data.frame(Variable   = names(imp_radial),
                     Importance = as.numeric(imp_radial))
imp_df <- imp_df[order(-imp_df$Importance), ]
readable_vars <- c(
  AGE       = "Age",
  SEX       = "Sex",
  POVERTY   = "Income-to-poverty ratio",
  BMICALC   = "BMI",
  MOD10DMIN = "Moderate exercise (min)",
  HRSLEEP   = "Sleep (hrs/night)",
  ALCDAYSYR = "Alcohol days/year",
  FRUTNO    = "Fruit servings/week"
)
imp_df$Variable <- readable_vars[as.character(imp_df$Variable)]

importance_plot <- ggplot(imp_df,
                          aes(x = reorder(Variable, Importance),
                              y = Importance)) +
  geom_col(fill = "#d62728") +
  coord_flip() +
  labs(title    = "Permutation Variable Importance (Radial SVM)",
       subtitle = "Drop in test AUC when each variable is shuffled",
       x = NULL, y = "AUC drop") +
  theme_minimal(base_size = 13)

print(importance_plot)
ggsave("variable_importance.png", importance_plot, width = 8, height = 5, dpi = 300)

print(imp_df)
```

```{r}
# Stop the parallel cluster
stopCluster(cl)
registerDoSEQ()
```

## Summary

Across all three kernels, the test AUC is around 0.78 and the differences between kernels are small. The permutation importance results show that AGE alone drives almost all of the predictive signal with every other variable contributing very little. This explains why the more flexible radial and polynomial kernels do not beat the linear kernel by much. The full discussion of these results, including limitations and implications is in the poster.
